Stability of variational quantum dynamics with Gaussian approximants

### 2013-06-18 paper.tex
write up changes from original plan

\title{The instability of variational quantum mechanics with Gaussian approximants is (not) entirely due to the unstable least squares problem}

@@

A numerical experiment was carried out to test this hypothesis.  The dynamics of a Morse oscillator were computed by three different methods.  First, a harmonic potential was fit to the bottom of the Morse potential, the state and the Hamiltonian were expanded over its eigenstates, and Schrödinger's equation was integrated directly in Hilbert space.  Secondly, differential equations were derived for the variational parameters in the standard way, using the normal equations of the regularised variational problem, and these were solved in parameter space.  The third method was intermediate: Schrödinger's equation was solved in Hilbert space, but the state was approximated by a superposition of coherent states, updated by least squares at each timestep.  Finally, the eigenstates of this potential are known exactly, and the state was expanded over those.

In a sense, expanding over Fock states is cheating.  If it were practical to do this, there would be no point using a variational method: these would be employed in many-particle problems, where an orthonormal basis would be unmanagably large.  In principle, the operations required for the second and third methods could be performed on the superpositions of coherent states, using the normal equations, without projection on an orthonormal basis; in a many-particle problem, they would have to be.  However, this would raise the question whether those least-square solutions from the normal equations were stable.  By using an orthonormal basis, we can be more confident of this.

A few different ODE solving algorithms were used, with different tolerance for stiff ODEs.  Adaptive 4th/5th order Runge-Kutta was used.  This is the algorithm that everyone will try by default, and it would be useful to know if it performs much worse than other algorithms.  ? and ? were used for comparison.  These discretisations have very different regions of stability ...

The first experiment computed the dynamics of a quartic oscillator, with Hamiltonian 
$$H=\hbar a^{2\dagger}a²,$$
solving Schrödinger's equation quite conventionally.  The quantum state at time $t$ was expanded over a truncated Fock basis $|N〉=\pmatrix{|1\rangle & |2\rangle & ⋯ & |N\rangle}$, with a column vector of coefficients $c(t)$, as
$$|ν(c(t))〉=|N〉c(t).$$
Schrödinger's equation expands to
$$|N〉Dc(t)=|D(ν\circ c)(t)〉={H\over i\hbar}|ν(c(t))〉=|N〉〈N|{H\over i\hbar}|N〉c(t),$$
whence
$$Dc(t)=-i〈N|{H\over \hbar}|N〉c(t).$$
Usually, these expansions would be approximate due to basis truncation.  In this case, the Fock states diagonalise $H$, so they are exact.  However, in general the initial state can only be approximately expanded over $|N\rangle$.  Time was discretised with the explicit Euler formula, which replaces the differential equation
$$Du(t)=f(u(t))$$
with a difference equation
$$v^{n+1}=vⁿ+τfⁿ,$$
where $tⁿ=nτ$, $vⁿ$ is meant to approximate $u(tⁿ)$, and $fⁿ=f(vⁿ)$.  Let
$$vⁿ=cⁿ\qquad{\rm and}\qquad fⁿ=-i〈N|{H\over\hbar}|N〉c(t).$$
Since $〈N|H|N〉$ is diagonal, the matrix product can be performed sparely as a dot product.  The results are shown below.  The Fock states are eigenstates of the Hamiltonian, so their amplitudes should be constant: in the exact solution of Schrödinger's equation, all the ratios plotted are exactly 1.

@@

A third method is the semi-implicit one, with the formulae
$$v^{\prime n}=vⁿ+½τf^{\prime n}\qquad v^{n+1}=vⁿ+τf^{\prime n}.$$
With the Hamiltonian, this becomes
$$c^{n+1}=cⁿ-iτH\biggl(1+{iτ\over 2}H\biggr)^{-1}cⁿ.$$

>>>>

\title{Stability of variational quantum dynamics with Gaussian approximants}

@@

A numerical experiment was carried out to test this hypothesis.  The dynamics of a quartic oscillator were computed by three different methods.  First, a harmonic potential was fit to the bottom of the Morse potential, the state and the Hamiltonian were expanded over its eigenstates, and Schrödinger's equation was integrated directly in Hilbert space.  Secondly, differential equations were derived for the variational parameters in the standard way, using the normal equations of the regularised variational problem, and these were solved in parameter space.  The third method was intermediate: Schrödinger's equation was solved in Hilbert space, but the state was approximated by a superposition of coherent states, updated by least squares at each timestep.  Finally, the eigenstates of this potential are known exactly: the state was expanded over them, and the exact dynamics calculated for comparison.

In a sense, expanding over Fock states is cheating.  If it were practical to do this, there would be no point using a variational method: these would be employed in many-particle problems, where an orthonormal basis such as the Fock states would be unmanagably large.  In principle, the operations required for the second and third methods could be performed on the superpositions of coherent states, using the normal equations, without projection on an orthonormal basis; in a many-particle problem, they would have to be.  However, this would raise the question whether those least-square solutions from the normal equations were stable.  By using an orthonormal basis, we can be more confident of this.

I used three discretisation formulae for ODE, with different regions of stability.  As a first step, the Hamiltonian and a coherent state with real amplitude 2 was expanded over Fock states $|0〉$ through $|10〉$, and the resulting equations for the components solved directly.  The Hamiltonian is
$$H=\hbar a^{2\dagger}a²,$$
so the Fock states are eigenstates, although the spacing of their energies is uneven.  The quantum state at time $t$ was expanded over a truncated Fock basis $|N〉=\pmatrix{|1\rangle & |2\rangle & ⋯ & |N\rangle}$, with a column vector of coefficients $c(t)$, as
$$|ν(c(t))〉=|N〉c(t).$$
Schrödinger's equation expands to
$$|N〉Dc(t)=|D(ν\circ c)(t)〉={H\over i\hbar}|ν(c(t))〉=|N〉〈N|{H\over i\hbar}|N〉c(t),$$
whence
$$Dc(t)=-i〈N|{H\over \hbar}|N〉c(t).$$
Usually, these expansions would be approximate due to basis truncation.  In this case, the Fock states diagonalise $H$, so they are exact.  However, in general the initial state can only be approximately expanded over $|N\rangle$.  Time was discretised with the explicit Euler formula, which replaces the differential equation
$$Du(t)=f(u(t))$$
with a difference equation
$$v^{n+1}=vⁿ+τfⁿ,$$
where $tⁿ=nτ$, $vⁿ$ is meant to approximate $u(tⁿ)$, and $fⁿ=f(vⁿ)$.  Let
$$vⁿ=cⁿ\qquad{\rm and}\qquad fⁿ=-i〈N|{H\over\hbar}|N〉c(t).$$
Since $〈N|H|N〉$ is diagonal, the matrix product can be performed sparely as a dot product.  The results are shown below.  The Fock states are eigenstates of the Hamiltonian, so their amplitudes should be constant: in the exact solution of Schrödinger's equation, all the ratios plotted are exactly 1.

@@

A third method is the semi-implicit one, with the formulae
$$v^{\prime n}=vⁿ+½τf^{\prime n}\qquad v^{n+1}=vⁿ+τf^{\prime n}.$$
With the Hamiltonian, this becomes
$$c^{n+1}=cⁿ-iτH\biggl(1+{iτ\over 2}H\biggr)^{-1}cⁿ.$$
If Hilbert space is considered as a real vector space, complex multiples of the eigenstates of the Hamiltonian form planes instead of lines.  The eigenstates evolve as $e^{iωt}$, so, when the dynamics are projected onto an eigenplane, the state traces out a circle, with radius equal to the absolute value of the the coefficient of the eigenstate when the initial state is expanded.  Different eigenstates circle the origin with different frequencies.

The semi-implicit formula has a special property: when the trajectory of $u(t)$ is a circle, the $vⁿ$ lie exactly on that circle, and the discretisation error is equivalent to rescaling time by a factor that depends on the frequency and the discrete timestep.  In Hilbert space, this means that the projections of the state onto eigenplanes still circle the origin exactly: only the frequencies are altered by discretisation.  So the effect of discretising quantum dynamics using the semi-implicit formula is merely to shift the energies of the eigenstates.  Of course, this can be a significant effect: it's the only difference between the harmonic and quartic oscillators.
